A More Complete Algebra for File Synchronization

### 2015-11-17 2015/paper.tex
New axioms and definitions

@@

The paths of a filesystem form a disjoint union of rooted directed trees.
\begin{mydef}
The function $\parent(p)$ returns the parent path of $p$, or
returns $\toppath$ if $p$ is the root of a tree.
We also write $\pp$ for $\parent(p)$.
\end{mydef}

@@

We therefore propose to have a command for each pair of types.
For want of a better system, we will name our commands by concatenating
two of $\fsregcommandchar{D}$, $\fsregcommandchar{F}$ and $\fsregcommandchar{\empt}$, 
where the left sign notes the type of value
in the filesystem before the command is applied, and the right sign notes the type
afterwards. For example $mkdir(p,D)$ is $\cbd(p,D)$ and $rmdir(p)$ is $\cdb(p)$.
Please find their exact definitions below.

@@

@@

\section{Sequences of Commands}

% -----------------------------

We aim to use our algebra to define and implement algorithms for conflict detection
and ordering updates (commands) before we can apply them to all replicas.
Conflicts and ordering all happen between commands, therefore in our algebra
we will reason about sequences of commands, 
and do so based on a set of axioms about pairs of neighbouring commands.
We are interested to see which pairs of commands cause errors all the time,
and which can be simplified or reversed
in order to see under what circumstances commands conflict or can be reordered.

Pairs of commands in general will have the form
\[ [\cxy(p,V_Y); \czw(q,V_W)] \]
where $X,Y,Z,W\in\{\fsregcommandchar{D},\fsregcommandchar{F},\empt\}$, $p,q\in\setp$, 
and values are of the appropriate type: 
$V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$.

Similarly to the case of commands, in \cite{NREC:alg} we found that in order to build
a set of axioms in a sound and complete algebra and use the algebra effectively we at times needed to take into account
the relationship between the paths of the commands in the pair, $p$ and $q$.
As the relationship between the paths determine how the commands affect the filesystem
and its tree property, 
and again in line with the aim of providing as much information as possible for the algebra,
we encode this relationship in the pair of commands.

The possible relationships we
consider are as follows.

% One may think that we may be able to specify more about the reversibility
% or possible simplification of a pair of commands in the last group if we also
% knew whether the child path $p$ was the only (non-empty) child of its parent.
% However, it is easy to see that encoding this information in a pair
% is not going to enrich our algebra as, for example,
% the behaviour of a hypothetical $[\cfb(p_{\mathrm{only\ child}}); \cdb(\pp)]$
% is exactly the same as that of the more generic
% $[\cfb(p); \cdb(\pp)]$. The former will break any filesystem in which
% $p$ is not the only child as it violates its preconditions, but so will
% the generic pair.

% TODO Would it be worth encoding non-direct parent/child relationships?

%% What does it mean to restrict the inspected filesystems to a given relationship?

@@

\begin{mydef}{Assertion command.}
A command is an assertion command if
for every filesystem it either breaks it or leaves it in the same state.
In our current algebra $\cbb$ and $\cdd$ are the only two assertion commands.
\end{mydef}

@@

@@

\begin{myax}\label{worksextpostfix}
$\worksc{A}{A;S}$ where $A$ and $S$ are arbitrary sequences
\end{myax}

@@

% ---------------------------------------------------------------

Finally, we have a set of axioms or laws that describe the behaviour
of each command pair. We exclude commands that either break the
filesystem or leave it in the same state ($\cbb$) and for
simplicity's sake instead of listing axioms for all pairs of commands
we describe the behaviour of groups of pairs
based on the relationship between the paths in the commands.

\begin{description}
\item[Unrelated.] If the two paths in the pair are not directly related, then
the commands commute with an equivalence and they will not break every filesystem:
\begin{align*}
[\cxy(p,V_Y); \czw(\np,V_W)] &\equiv [\czw(\np,V_W); \cxy(p,V_Y)]
\end{align*}
%
\item[Same.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), the
the commands cannot be reversed. Instead:
   \begin{description}
   \item[Break.]
   The pair will break every filesystem if $Z\ne Y$, that is, if the second
   command expects a type of value different from the one left by the first command.
   \item[Simplified by extension.]
   Otherwise, the pair will simplify to a single or no commands
   if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
   \begin{gather*}
   % bf=fb Sp
            [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
   % bd=db Sp
            [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
   % db=bd Sp
            [\cdb(p); \cbd(p, D)] \eqext [\cdd(p, D)] \\
   % df=fd Sp
            [\cdf(p, F); \cfd(p, D)] \eqext [\cdd(p, D)]
   \end{gather*}
   \item[Simplified by equivalence.]
   In all other cases, the commands are equivalent to a single command
   replacing the input of the first command with the output of the second:
   \[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
   \end{description}
%
\item[Down.]
These are the pairs in which the first command is applied to the parent of the path
of the second command.
   \begin{description}
   \item[Construction.] There are pairs in this group to which no law applies:
   the commands do not commute or break every filesystem, and cannot be simplified.
   We call these ``construction'' pairs:
   \begin{gather*}
   % bd>bf
            [\cbd(\pp, D); \cbf(p, F)] \\
   % bd>bd
            [\cbd(\pp, D_1); \cbd(p, D_2)] \\
   % fd>bf
            [\cfd(\pp, D); \cbf(p, F)] \\
   % fd>bd
            [\cfd(\pp, D_1); \cbd(p, D_2)]
   \end{gather*}
   \item[Commute.] Otherwise, the commands in these pairs commute iff the first command is $\cdd$. For example:
   \begin{gather*}
   [\cdd(\pp, D); \cbf(p, F)] \equiv [\cbf(p, F); \cdd(\pp, D)]
   \end{gather*}
   \item[Break.] All other pairs break every filesystem, e.g.:
   \begin{gather*}
   [\cdf(\pp, F); \cbd(p, D)] \equiv \cbrk
   \end{gather*}
   \end{description}
%
\item[Up.]
The final group of pairs are the ones in which the path in the second
command is the parent of the one in the first.
These behave similarly to the previous group. 
   \begin{description}
   \item[Desctruction.] There are
   so-called ``destruction'' pairs to which no law applies:
   \begin{gather*}
   % fb<db ??
            [\cfb(p); \cdb(\pp)] \\
   % fb<df ??
            [\cfb(p); \cdf(\pp, F)] \\
   % db<db ??
            [\cdb(p); \cdb(\pp)] \\
   % db<df ??
            [\cdb(p); \cdf(\pp, F)]
   \end{gather*}
   \item[Commute.] Otherwise the commands commute iff the \emph{second} command is $\cdd$.
   \item[Break.] All other pairs break every filesystem.
   \end{description}
\end{description}

\section{Real-Life Variant}

In real-life applications, we often do not want to consider metadata stored in
directories (\cite{BZ}). In other words, the contents of directories are all equal,
which can be modelled by assuming that $|\setd|=1$.
Note that this creates a convenient symmetry where there is only a single value
for each type that can be repeated as we move from a path to its child ($D$ and $\empt$).

This also means that $\cdd$, like $\cbb$, becomes a command that either breaks a filesystem
or leaves it in the same state, and we will disregard it in our algebra.
The axioms will simplify as follows:

\begin{description}
\item[Unrelated.] Pairs with unrelated paths are still reversible and not equivalent to
$\cbrk$, but these are the only pairs that are reversible.
%
\item[Same.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then,
as above, the commands do not commute.
   \begin{description}
   \item[Break.] Also, as above, the pair will break every filesystem if $Z\ne Y$.
   \item[Simplified by extension.] Otherwise, the pair is extended by an empty sequence
   if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
   \begin{gather*}
   % bf=fb Sp
            [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
   % bd=db Sp
            [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
   % db=bd Sp
            [\cdb(p); \cbd(p, D)] \eqext \emptyseq \\
   % df=fd Sp
            [\cdf(p, F); \cfd(p, D)] \eqext \emptyseq
   \end{gather*}
   \item[Simplified by equivalence.]
   And in all other cases, as above, the commands are equivalent to a single command
   replacing the input of the first command with the output of the second:
   \[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
   \end{description}
%
\item[Down.]
Of the pairs in which the first command is applied to the parent path:
   \begin{description}
   \item[Construction.] the same ``construction'' pairs still have no laws, 
   \item[Break.] but all other pairs break all filesystems.
   \end{description}
%
\item[Up.]
And of the pairs where the second command
is applied to the parent path,
   \begin{description}
   \item[Destruction.] we still have the ``destruction'' pairs,
   \item[Break.] and all other pairs break all filesystems.
   \end{description}
\end{description}

% TODO
{\bf Going forward we use the algebra derived from the $|\setd|=1$ case.}

@@

We reintroduce two assertion commands that either break the filesystem
or leave it unchanged: $\cbb(p)$ and $\cdd(p)$.
(As $|\setb|=|\setd|=1$, we do not note their exact output value.)
They can be used to test whether $\FS(p)=\empt$ or $\FS(p)=D$.

Furthermore, we note the following axioms:

>>>>

\newcommand{\typeset}{\{\fsregcommandchar{D},\fsregcommandchar{F},\empt\}}

\newcommand{\descendant}{\prec}
\newcommand{\ancestor}{\succ}

@@

The paths of a filesystem form a disjoint union of rooted directed trees.
\begin{mydef}
The function $\parent(p)$ returns the parent path of $p$, or
returns $\toppath$ if $p$ is the root of a tree.
\end{mydef}

@@

We therefore propose to have a command for each pair of types.
For want of a better system, we will name our commands by concatenating
two of $\fsregcommandchar{D}$, $\fsregcommandchar{F}$ and $\fsregcommandchar{\empt}$, 
where the left sign notes the type of value
in the filesystem before the command is applied, and the right sign notes the type
afterwards. For example $mkdir(p,D)$ is $\cbd(p,D)$ and $rmdir(p)$ is $\cdb(p)$.

When we describe multiple or unknown commands, we may substitute one or both
types with a variable, e.g. $\cxy(p,V)$. In this context,
if we write that $\cxy=\czw$, we mean that their input and output types
are the same ($X=Z$ and $Y=W$).

@@

% TODO Only one directory value
In real-life applications, we often do not want to consider metadata stored in
directories (\cite{BZ}). In other words, the contents of directories are all equal,
which can be modelled by assuming that $|\setd|=1$.
Note that this creates a convenient symmetry where there is only a single value
for each type that can be repeated as we move from a path to its child ($D$ and $\empt$).

@@

\section{Notation and definitions}

\begin{mydef}{$p\descendant q$.}
$p\descendant q$ if $p$ is the ancestor of $q$ ($p=\parent^n(q)$).
\end{mydef}

\begin{mydef}{Assertion command.}
A command is an assertion command if
for every filesystem it either breaks it or leaves it in the same state.
In other words, assertion commands have the same input and output type,
and there must also be only one possible value for their output.
Accordingly in our current algebra $\cbb$ and $\cdd$ are the only two assertion commands.
As $|\setb|=|\setd|=1$, we do not note their exact output value.
\end{mydef}

@@

@@

% TODO A=B -> B=A; A=B B=C -> A=C; etc.

@@

@@

We aim to use our algebra to define and implement algorithms for conflict detection
and ordering commands before we can apply them to all replicas.
Conflicts and ordering all happen between commands, therefore in our algebra
we will reason about sequences of commands, 
and do so based on a set of axioms about pairs of neighbouring commands.
We are interested to see which pairs of commands cause errors all the time,
and which can be simplified or reversed
in order to see under what circumstances commands conflict or can be reordered.

Pairs of commands in general will have the form
\[ [\cxy(p,V_Y); \czw(q,V_W)] \]
where $X,Y,Z,W\in\typeset$, $p,q\in\setp$, 
and values are of the appropriate type: 
$V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$.

The following axioms describe multiple statements
that we accept as true in the algebra.
These statements are generated by substituting 
all possible values for
$X,Y,Z,W\in\typeset$; $p,q\in\setp$; $V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$
under the conditions noted in the axiom.

\begin{myax}\label{ax_separate_commute}
$\cxy(p,V_Y); \czw(q,V_W) \equiv \czw(q,V_W); \cxy(p,V_Y)$ where $p\not\descendant q$ and $q\not\descendant p$.
\end{myax}

\begin{myax}\label{ax_separate_nobreaks}
$\cxy(p,V_Y); \czw(q,V_W) \nequiv \cbrk$ where $p\not\descendant q$ and $q\not\descendant p$.
\end{myax}

\begin{myax}\label{ax_same_breaks}
$\cxy(p,V_Y); \czw(p,V_W) \equiv \cbrk$ if $Y\ne Z$.
\end{myax}

\begin{myax}\label{ax_distantchild_breaks}
$\cxy(p,V_Y); \czw(q,V_W) \equiv \cbrk$ if $p\descendant q$ and $p\neq\parent(q)$ and $\cxy\neq\cdd$ and $\czw\neq\cbb$.
\end{myax}

\begin{myax}\label{ax_distantparent_breaks}
$\cxy(p,V_Y); \czw(q,V_W) \equiv \cbrk$ if $p\ancestor q$ and $\parent(p)\neq q$ and $\cxy\neq\cbb$ and $\czw\neq\cdd$.
\end{myax}

\begin{mydef}{Construction pair.}
A pair of commands on paths $p$ and $q$ is a construction pair if $p=\parent(q)$ and the commands are one
of the following:
   \begin{gather*}
   % bd>bf
            [\cbd(p, D); \cbf(q, F)] \\
   % bd>bd
            [\cbd(p, D_1); \cbd(q, D_2)] \\
   % fd>bf
            [\cfd(p, D); \cbf(q, F)] \\
   % fd>bd
            [\cfd(p, D_1); \cbd(q, D_2)]
   \end{gather*}
\end{mydef}

\begin{mydef}{Destruction pair.}
A pair of commands on paths $p$ and $q$ is a destruction pair if $\parent(p)=q$ and the commands are one
of the following:
   \begin{gather*}
   % fb<db ??
            [\cfb(p); \cdb(\pp)] \\
   % fb<df ??
            [\cfb(p); \cdf(\pp, F)] \\
   % db<db ??
            [\cdb(p); \cdb(\pp)] \\
   % db<df ??
            [\cdb(p); \cdf(\pp, F)]
   \end{gather*}
\end{mydef}

\begin{myax}\label{ax_directchild_breaks}
$\cxy(p,V_Y); \czw(q,V_W) \equiv \cbrk$ if $p=\parent(q)$ and $\cxy\neq\cdd$ and $\czw\neq\cbb$
and the pair is not a construction pair.
\end{myax}

\begin{myax}\label{ax_directparent_breaks}
$\cxy(p,V_Y); \czw(q,V_W) \equiv \cbrk$ if $\parent(p)=q$ and $\cxy\neq\cbb$ and $\czw\neq\cdd$
and the pair is not a destruction pair.
\end{myax}

\begin{myax}\label{worksextpostfix}
$\worksc{A}{A;S}$ where $A$ and $S$ are arbitrary sequences of commands
\end{myax}

% Separate:
% * commute
% - if both are BB or DD, extended by empty seq
% - if either is BB or DD, extended by the other command
% - not equiv to break
% Same:
% * equiv break unless 1-out == 2-in
% - extended by empty seq if 1-out==2-in & 1-in==2-out==B,D
% - extended by single command (if 1-out==2-in) (otherwise breaks so true)

@@

We note the following axioms:
